\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-01-05 writing/manuscript.Rmd
read up to section on modeling HR

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

@@

\noindent Since animals need to maintain a positive energetic balance to grow and reproduce, their spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance in around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resource abundance on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. And while abundant work has been done on estimating animal's response to changes in resource abundance, there is little to no information on how many species respond to unpredictable changes in the resources they depend on.

A habitat's long-term average resource abundance is not sufficient to assess the habitat's quality. A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted. Although both habitats may have the same long-term resource abundance, the differences in resource predictability result in substantially different habitats. Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].

In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive chapters. Concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology. Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis of how animals respond to changes in resource abundance. Afterwards, we present a hypothesis for how animals respond to resource unpredictability. Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

@@

\noindent In probability and statistics, random variables are defined by specifying the distribution the variable follows. Since the variable is random, it can take multiple values (minimum 2), each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.

@@

\noindent The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.

@@

\noindent A random variable's variance is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$ (We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is $\mu = \frac{\alpha}{\alpha+\beta}$ while the variance is $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$). Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

@@

\noindent @nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

\begin{equation} \label{nathan-movement}
\vec u_{t+1} = F(\Omega, \Phi, \vec r_t, \vec w_t, \vec u_t),
\end{equation}

such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple. However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly).

@@

\noindent To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

If sampling frequency is too low to provide precise measures of autocorrelation, it will be inappropriate to estimate an animal's average speed, directional persistence, and range crossing time, but the data can still provide reasonable estimates of the animal's spatial needs and usage, and model selection procedures with criteria such as AIC can help the analyst choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022]. Estimates of average or instantaneous speed require the highest sampling frequency [$\Delta t \lessapprox \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction. However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior, it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed. Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the sampling duration is sufficiently long.

@@

\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident. ***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***

Although range-resident animals will generally have smaller $\vpos$ than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior. Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement. Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.

### Advantages of modeling animal movement as a stochastic process

\noindent Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

Depending on sampling frequency, statistical methods that rely on straight-line displacement are prone to under-estimating the distance traveled by an animal and its speed at coarse sampling frequencies, while they over-estimate them at fine frequencies [@noonan_scale-insensitive_2019]. In addition, most commonly-used methods for home range estimation assume tracking data to be independent of one another and sampled at consistent and discrete intervals, which is most often not the case. Consequently, animals' spatial needs are often under-estimated [@noonan_comprehensive_2019].

### Unassigned

@pretorius_movement_2020: flamingos -- migrating or nomadic?

@boratynski_energetic_2020: animals' hr is limited by BMI and maximum metabolic rate, which can be accounted for by accounting for body mass [@noonan_effects_2020]

@arechavala-lopez_common_2019: octopuses (*Octopus vulgaris*) congregate in human-altered coastal habitats

@mitsuhashi_home_2018: urban raccoon dogs (*Nyctereutes procyonoides*) have a smaller HR than rural ones

@cronk_home_2021: yellow mongoose (*Cynictis penicillata*) in urban regions have smaller HRs

@yu_artificial_2022: rat snakes move to artificial structures for better prey availability and better thermoregulation

Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].

@teitelbaum_beyond_2019:

- **Nomadic movement**: movement with irregular timing and/or direction that produces both within- and between-year variability in location and movement patterns
- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north Atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"

@@

# Effects of $\e(R)$

@@

The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.

competition can promote niche partitioning [@fox_niche_1981]

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    # - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats     # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

@@

\noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], and reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @ref; mammals: @ref], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home range size is

\begin{equation} \label{harestad-eq}
H = C / R,
\end{equation}

\noindent where $H$ is the animal's **home range size**, $C$ is the animal's resource consumption, and $R$ is the abundance of (usable) resources. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. And while abundant work has been done on estimating animals' responses to changes in resource abundance, there is little to no information on they respond to unpredictable changes in the resources they depend on.

In this paper, we refer to a location's average amount of available and accessible resources as **resource abundance**, while use the phrase **resource unpredictability** to indicate the variability in resource abundance due to *unpredictable* causes. On its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality. A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted. Although both habitats may have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different habitats. Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].

In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections, and concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology. Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource unpredictability on animals spatial needs while suggesting a second unifying hypothesis. Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

@@

\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable follows. Since the variable is random, it can take multiple values (minimum 2), each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.

The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers (i.e., from $-\infty$ to $\infty$), while the Bernoulli distribution only has support over values 0 and 1.

@@

\noindent The **expectation** of a random variable (i.e. its mean or average, indicated here with the symbol $\e$) is the value one can *expect* the random variable to take. $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$. We indicate the estimated average resource abundance by adding a caret on the symbol: $\hat \mu(t)$.

@@

\noindent A random variable's **variance** is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$, and $\widehat{\sigma^2}(t)$ to indicate its estimate conditional on time. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

\begin{equation} \label{beta-mean}
\mu = \frac{\alpha}{\alpha + \beta}
\end{equation}

\noindent while the variance is

\begin{equation} \label{beta-var}
\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
\end{equation}

\noindent Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

@@

\noindent @nathan_movement_2008 generalized animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model

\begin{equation} \label{nathan-movement}
\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
\end{equation}

\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple. Thus, we can instead focus on animals' behavior and spatial needs as a response to overall resource abundance and unpredictability.

@@

\noindent To understand how movement can be viewed as a **stochastic process**, imagine an animal that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position). This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.

By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data. While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].

Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction. However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed. Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the total **sampling duration**, $T$, is sufficiently long. Model selection procedures with criteria such as AIC [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less *why* it moved. Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident. In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs using a sliding window approach (see the discussion section).

**I don't think we need this paragraph:** Although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior. Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors. Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have a position it returns to regularly.

@@

# Effects of resource abundance, $\e(R)$

@@
